\chapter{Motivation and Context}\label{chap:introduction}

In \citeyear{rontgen_uber_1895} \citeauthor{rontgen_uber_1895} discovered X-rays. Since then
different methods based on X-ray have been revolutionizing medical treatment. For the first time it
was possible to inspect the human body non-intrusively, i.e.\ without cutting it open. It was used
among others to detect metal objects in the body such as bullets~\cite{haygood_skeletal_1996}. But
also new kinds of fractures could be detected and diagnosed. New bone diseases such as
Osteopoikilosis could be discovered, and others such as skeletal
tuberculosis~\cite{haygood_skeletal_1996} could be identified.

The next major step was the invention of \textit{X-ray attenuation computed tomography} (CT). The
first prototypes were developed at EMI laboratories and resulted in a series of publications
~\cite{hounsfield_computerized_1973,ambrose_computerized_1973,perry_computerized_1973}. Compared to
simple X-ray images, X-ray attenuation CT can create cross-sectional images from the interior of the
scanned object. Whereas, in tradition X-ray imaging different objects superimpose each other, X-ray
attenuation CT can distinguish them. Since its introduction, X-ray attenuation CT became a standard
tool in medical diagnostics. Medical applications include, but are by no means limited to, the
detection of infarction, tumors, calcifications, hemorrhage and bone
trauma~\cite[Chapter~5]{buchanan_advanced_2012}. However, X-ray attenuation does lack soft-tissue
contrast~\cite{pfeiffer_phase_2006}. I.e.\ tissues with similar attenuation coefficient are hard to
distinguish.

Phase-contrast X-ray imaging improves on this specific issues~\cite{lewis_medical_2004}. Plus it can
reveal additional information~\cite{hahn_numerical_2012}. One way to measure phase-contrast for
X-rays is grating interferometer~\cite{pfeiffer_hard-x-ray_2008}, often referred to as
\textit{grating-based}. Phase information can be indirectly measured, via the refraction of X-rays,
instead of the attenuation. Specifically, the phase gradient perpendicular to the gratings can be
measured, therefore, this method is often called \textit{differential phase-contrast imaging}.

Additionally, the same grating-based setup yields further information about (ultra) small-angle
scattering. Imaging modalities based on this property are referred to as \textit{X-ray dark-field
	imaging}. Studies have been proposed for numerous different medical applications,
see~\cite[Chapter 1.3.1]{wieczorek_anisotropic_2017} for an overview of the different fields. X-ray
dark-field imaging is still a field of active research. Only recently, a system was proposed, which
incorporated such a grating-based setup into a clinical CT scanner~\cite{viermetz_dark-field_2022}.

This is a part of the working context of the research group \textit{\gls{CIIP}} at the \gls{TUM}
around my supervisor PD Dr.\ rer.\ nat.\ Tobias Lasser. In particular, the open-source C++ framework
\textit{elsa} is developed at the research group. Its main use is in the field of tomographic
reconstruction. One goal of elsa is to provide open access to the research done at the group, plus
it should improve the reproducibility of the work done. Therefore, support for imaging modalities
such as phase-contrast or dark-field imaging enables the research group to increase the
accessibility and reproducibility of their research. But --- like many other known frameworks ---
elsa currently only supports X-ray attenuation CT\@. Due to the aforementioned accessibility and
reproducibility reasons, there exists high interest to support both X-ray phase-contrast and
dark-field imaging modalities in elsa. This support should include common, state-of-the-art and
novel algorithms for tomographic reconstruction. Without common and current state-of-the-art
algorithms, no proper evaluation and/or comparison to novel techniques can be made.

Though, they are different modalities, they do have a common basis. All of the above mentioned
imaging modalities can be sorted in the group of \textit{tomographic reconstruction}, which itself
is a subset of \textit{inverse problems}. The challenge for this specific class of problems is the
reconstruction of a signal from a finite set of projections~\cite{herman_basis_2015}. The connection
between the problem domain and the measurement domain is the so-called \textit{forward model}.

The forward model for X-ray attenuation CT, differential phase-contrast CT and dark-field CT is
based on the line integral. They assume X-rays traveling throughout some object in a straight line.
Hence, some quantity (i.e.\ the attenuation coefficients of an object) can be integrated along the
path of the X-ray. Such a quantity can be fully described by the infinite set of lines from
different projection angles. This has been proven multiple times. However, most often it is
attributed to \citeauthor*{radon_uber_1917} in \citeyear{radon_uber_1917} ~\cite{radon_uber_1917}.
Therefore, the integral transform, which maps a function into its set of line integrals is referred
to as the \textit{Radon Transform}. Other than \citeauthor*{radon_uber_1917},
\citeauthor*{cormack_representation_1963}~\cite{cormack_representation_1963} rediscovered it without
the knowledge of the work done by \citeauthor*{radon_uber_1917}.
\citeauthor*{cormack_representation_1963} laid the theoretical foundation of modern CT scanners. And
\citeauthor*{hounsfield_computerized_1973}~\cite{hounsfield_computerized_1973} built the first
working prototypes. For that, both \citeauthor*{cormack_representation_1963} and
\citeauthor*{hounsfield_computerized_1973} received the Nobel Prize in Medicine in 1979.

To successfully reconstruct a signal from its projections, a couple of key steps need to be
performed. First, the forward model has to be described precisely. Next, the problem needs to be
discretized. This leads to a system of linear equations, which needs to be solved as the final step.

Of great interest for this thesis is the discretization aspect. A common approach to discretization
is the series expansion~\cite{herman_basis_2015}. The signal is described via a coefficient vector
and a set of basis functions. The choice of basis function has great influence on the quality of the
discretization. But not only that, usually signals in tomographic reconstructions are considered
smooth, i.e.\ continuously differentiable. However, arguably the most common basis functions --- the
pixel or voxel basis function --- is not differentiable at its boundaries. This creates problems in
imaging modalities that expect a differentiable problem domain, such as differentiable
phase-contrast imaging. Therefore, it is of interest to investigate basis functions, which are
continuously differentiable. Precisely this study of differentiable basis functions is an important
topic of this thesis.

\citeauthor*{lewitt_multidimensional_1990} introduced a family of spherically-symmetric basis
functions often referred to as \textit{blobs} in~\cite{lewitt_multidimensional_1990}. It received a
lot of attention in the tomographic space, as it overcomes quite a few limitations of the pixel and
voxel basis functions. Among others, it is continuously differentiable. Another interesting set of
basis functions is not as common in the tomographic space: Basis-Splines (or usually only
B-Splines). B-Splines are mostly known from function approximation. There exists a closed form
solution for a continuous approximation of signals using B-Splines~\cite{unser_fast_1991}. This
thesis follows the approach first laid out by \citeauthor*{momey_new_2011} in \cite{momey_new_2011}.
A lot of the properties of B-Splines are still relevant, as e.g.\ they tend to spherically symmetric
functions, but preserve a compact support. Further, they are --- up to a certain degree ---
continuously differential.

Once the problem domain is discretized using another basis function apart from the pixel or voxel
basis function, another issue needs to be solved. A considerable amount of methods exist to compute
the line integral. However, most methods assume the pixel or voxel basis function. Hence, they have
to be extended to use other basis functions.

The last two points are the main contribution of this thesis. The approximation of the line integral
relying on differentiable basis functions rather than the pixel or voxel basis functions are
implemented in elsa. This implicitly includes a discretization of the signal in the new basis
functions. However, it should be noted, that at no point in time the problem domain is transformed
to a new basis function.

\section{Outline}\label{sec:outline}

As already mentioned, there are three steps involved to tomographic reconstruction on computers.
There needs to be an understanding and description of the forward model. The problem needs to be
discretized, and finally a solution needs to be computed. These parts are covered in
\autoref{part:foundation}

The discussion about the forward model is split into two parts. The first, given in
\autoref{chap:imaging_modalities}, starts with a short introduction of the mathematical class of
inverse problems followed by a description of X-ray attenuation CT\@ and differential phase-contrast
CT\@. This includes an introduction to the physical and mechanical mechanism of the different
methods. Overall, this section should present an --- hopefully --- approachable, but engineering
focused intuition for tomographic reconstruction.

The second part of the discussion about the forward model, is the mathematical basic forward model
for tomographic reconstruction. It is described in \autoref{chap:radon_transform_and_related}. There,
important mathematical concepts, such as the Radon Transform, are introduced and discussed.

With that, the first of the three steps is discussed. In \autoref{chap:signal_representation}, the
discretization of the problem is discussed in detail. This includes the discussion of the pixel or
voxel basis function, the family of spherically-symmetric basis function introduced by
\citeauthor*{lewitt_multidimensional_1990} and B-Splines basis functions.

As a final piece of the puzzle, an introduction to finding a solution to tomographic reconstruction
problem is given in \autoref{chap:tomographic_reconstruction}. This includes a discussion of
analytical, iterative and regularized reconstruction algorithms.

Moving away from a mathematical centered discussion, \autoref{part:practical} starts with a short
introduction to the C++ framework elsa in \autoref{chap:elsa}. This is followed by a detailed
discussion of aspects regarding the implementation of computation of the line integral, with a focus
on other basis functions in \autoref{chap:projector}. Finally, by means of experiments, a detailed
evaluation of the new implementations of the line integral is conducted in
\autoref{chap:experiments}.

\autoref{part:conclusion} completes the thesis with a summary of the newly implemented projectors in
\autoref{chap:summary}, and an outlook for future work in \autoref{chap:future_work}.

\section{Disclaimer}\label{sec:disclaimer}

I want my research to be as open, reproducible and comprehensible as possible. Hence, the
development of the source code I have written is open source, discussions (though most of the
discussion was not online) can be found in the \href{https://gitlab.lrz.de/IP/elsa}{elsa GitLab
	repository}. The \LaTeX{} source code of thesis can be found on my
\href{https://github.com/ner0-m/ma-thesis-differential-basis-for-ct}{personal GitHub page}. To the
best of my possibilities, I try to include the code I have used to generate all of my plots and
graphs, I try to state dependencies and make building and using my work as easy as I can. But of
course, I am aware, what is easy to me, might be cumbersome to the next, and impossible to the
other.

Throughout the thesis, I specifically try to not mention any form of gender specific pronouns.
Instead, mostly \textit{one} or sometimes \textit{they} is used instead. Wherever relevant, I try to
use the gender-neutral form of wording to include all readers of whatever background. Please note
though, that this is a best effort. Not only is English not my first language, but also the
background, I grew up in and still live, and the socialization I underwent, are dominated by a
mostly gender binary form of thinking. Although, I oppose these views and do my best to educate
myself, I cannot deny, they influence my thinking and writing.
