\thispagestyle{empty}
\chapter*{Acknowledgements}
\noindent%

Firstly, of course, I have to thank my supervisor Tobias Lasser! We have been working together now
for more than 5 years. It all started with a Bachelor seminar, turned into my Bachelor's thesis,
followed by many years of research assistant positions.

During the last years of my studies, I found myself struggling with the pressure and stress. Not
only during that time, but in general, Tobias always supported me in many ways, writing letters of
recommendation, providing me an amazing opportunity to do an internship in Denmark (shootout to Per
Christian Hansen and Hans Henrik Sørensen, both are amazing people), providing me access to
facilities or just show interest in weird my interest.

I sincerely thank Tobias, for the freedom he offered me. During all the different projects we worked
on, I never had the feeling of doing something, I was forced to and did not like myself. Rather, it
was the other way around. I could come up with esoteric ideas and given enough reason, or enough
persistence, I could spend time on it and explore. This is and was a great way to grow for me
personally. I never got the feeling of Tobias being my superior, but rather I felt treated as equal.
I've worked in a couple of industry jobs during the years, and though I learned a ton there as well,
I always longed for this exact feeling and came back. And I can't thank Tobias enough for that.

Then I also want to thank my family. I want to thank my parents. They supported me in so many ways,
throughout my studies. Of course, they provided me with the chances to my education and supported me
financially. But they also supported me emotionally.

I want to thank my brother, he is quite a bit older than I am, but throughout my youth, I loved
watching him play video games. This and his geekiness for computers and tech shaped many of my
interests, I still have till today. Precisely those interested brought me to study computer science
in the first place.

Big thanks to my sister. We grew up together and lived in a room for the best part of our live. But
during that time, we've laughed so so much. Till today, we share much of our humor and I don't think
that there is anyone out there, who understands my humor better than her. None can understand my
though jumps just quite like her. It took me a couple of years to understand, that that is not
self-evident, oh well.

Lastly, I want to thank - in no particular order - Magi, Luzi, Mira, Eli, Anna, Phillip, Alex,
Alaya, Magda, Pil, Storm, Nana, Anabel, Luco, Virgina, Richy for everything. For all the
discussions, late night talks, walks, breakfast talks, coffee breaks. For all the emotional support,
for putting up with me in all my different moods, for providing distraction from my moods. But most
of all, you lots shaped me the most in the last years. Without you, I wouldn't be the person, I am
today.

Without all of you, I would have not made it.

\begin{flushright}
	Thank you so so much.
\end{flushright}
